\begin{thebibliography}{99}

\bibitem{Aassila}
\textsc{Aassila} (Mohammed), \emph{1000 challenges mathématiques : Algèbre}, Paris, Ellipses, 2016.

\bibitem{thebook}
  \textsc{Aigner} (Martin) et \textsc{Ziegler} (G\"{u}nter M.), \emph{Raisonnements divins} (trad. N. Puech), Cachan, Hermès-Lavoisier 2017.

\bibitem{alperin}
  \textsc{Alperin} (Roger C.), \og A mathematical theory of origami constructions and numbers\fg, \emph{New York Journal of Mathematics}, \no 6, 2000, p. 119-133.

\bibitem{Bacaer}
  \textsc{Bacaër} (Nicolas), \emph{Histoires de mathématiques et de populations}, Paris, Cassini, 2008. \url{http://www.ummisco.ird.fr/perso/bacaer/}.

\bibitem{Bacaer2}
  \textsc{Bacaër} (Nicolas), \emph{Mathématiques et épidémies}, Paris, Cassini, 2021. \url{http://www.ummisco.ird.fr/perso/bacaer/}.


\bibitem{marita}
  \textsc{Barabash} (Marita), \og A non-visual counterexample in elementary geometry\fg, \emph{The College Mathematics Journal}, \no 36, 2005.

\bibitem{mlcs}
  \textsc{Ben-Ari} (Mordechai), \emph{Mathematical Logic for Computer Science}, Springer, 2012.


\bibitem{joss}
  \textsc{Ben-Ari} (Mordechai), \og {LearnSAT}: A {SAT} Solver for Education\fg, \emph{Journal of Open Source Software}, \no 24, 2018, p. 639, \url{https://doi.org/10.21105/joss.00639}.


\bibitem{jorg}
  \textsc{Bewersdorff} (J\"{o}rg), \emph{Galois Theory for Beginners: A Historical Perspective}, American Mathematical Society, 2006.

\bibitem{bold}
  \textsc{Bold} (Benjamin), \emph{Famous Problems of Mathematics: A History of Constructions with Straight Edge and Compass}, Van Nostrand, 1969.

\bibitem{bradford}
  \textsc{Bradford} (Phillips Verner), \og Visualizing solutions to n-th degree algebraic equations using right-angle geometric paths\fg, 2010, \url{https://web.archive.org/web/20100502013959/http://www.concentric.net/~pvb/ALG/rightpaths.html}.


\bibitem{burton}
  \textsc{Butler IV} (Lane), \og Ramsey Theory\fg, 2016, \url{https://www.whitman.edu/Documents/Academics/Mathematics/2016/Barton.pdf}.

\bibitem{callagy}
  \textsc{Callagy} (James J.), \og The central angle of the regular 17-gon\fg, \emph{The Mathematical Gazette}, \no 67, 1983, p. 290-292, \url{https://www.jstor.org/stable/3617271}.

\bibitem{Carrega}
\textsc{Carrega} (Jean-Claude), \emph{Théorie des corps: La règle et le compas}, Paris, Hermann, 1981.

\bibitem{courant}
  \textsc{Courant} (Richard) et \textsc{Robbins} (Herbert), \emph{Qu'est-ce que les mathématiques?} (trad. M. Anglade, K. Py), Paris, Cassini, 2015.


\bibitem{davies}
  \textsc{Davies} (Roy O.), \og On {Langford's} problem {(II)}\fg, \emph{The Mathematical Gazette}, \no 43, 1959, p. 253-255.

\bibitem{Debray}
\textsc{Debray} (Régis), \emph{Civilisation: Comment nous sommes devenus américains}, Gallimard, 2017.

\bibitem{Delahaye}
\textsc{Delahaye} (Jean-Paul), \emph{Les mathématiciens se plient au jeu: 20 enquêtes ludiques du poker à l'origami}, Paris, Belin, 2017. 

\bibitem{dorrie1}
  \textsc{D\"{o}rrie} (Heinrich), \emph{100 Problems of Elementary Mathematics: Their History and Solution}, Dover, 1965.


\bibitem{dorrie2}
  \textsc{D\"{o}rrie} (Heinrich), \og 100 Problems of Elementary Mathematics: Their History and Solution\fg, 2010, \url{https://web.archive.org/web/20191223032114/http://www2.washjeff.edu/users/mwoltermann/Dorrie/DorrieContents.htm}.


\bibitem{dudley-budget}
  \textsc{Dudley} (Underwood), \emph{A Budget of Trisections}, Springer, 1987.


%\bibitem{dudley-what}
 % \textsc{Dudley} (Underwood), \og What To Do When the Trisector Comes\fg, \emph{The Mathematical Intelligencer}, \no 5, 1983, p. 20-25.


\bibitem{eppstein}
  \textsc{Eppstein} (David), \og Twenty Proofs of {Euler's} Formula: {$V-E+F=2$}\fg, \url{https://www.ics.uci.edu/~eppstein/junkyard/euler/}.

\bibitem{euclid}
\textsc{Euclide}, \emph{Les \'Eléments} (trad. B. Vitrac),  Paris, Presses universitaires de France, 1990-2001.


\bibitem{fraleigh}
  \textsc{Fraleigh} (John B.), \emph{A First Course in Abstract Algebra}, Addison-Wesley, 2003.

\bibitem{gauss}
  \textsc{Gauss} (Carl Friedrich), \emph{Recherches arithmétiques} (trad. A. Poullet-Delisle), Sceaux, Jacques Gabay, 1989.


 
\bibitem{gelfand}
  \textsc{Gelfand} (I. M.) et \textsc{Saul} (Mark), \emph{Trigonometry}, Boston, Birkhäuser, 2001.

\bibitem{Gowers}
\textsc{Gowers} (Timothy), \emph{Combinatoire}, Collège de France / Fayard, 2021.

\bibitem{rudiments}
  \textsc{Graham} (Ron) et \textsc{Butler} (Steve), \emph{Rudiments of Ramsey Theory}, American Mathematical Society, 2015.


\bibitem{gunderson}
  \textsc{Gunderson} (David S.), \emph{Handbook of Mathematical Induction: Theory and Applications}, Mathematical Association of America, 2010.

\bibitem{brute}
  \textsc{Heule} (Marijn J. H.) et \textsc{Kullmann} (Oliver), \og The Science of Brute Force\fg, \emph{Communications of the ACM}, \no 60, 2017, p. 70-79.


\bibitem{hull-beloch}
  \textsc{Hull} (Thomas C.), \og Solving Cubics With Creases: The Work of {Beloch} and {Lill}\fg, \emph{American Mathematical Monthly}, \no 118, 2011, p. 307-315.

\bibitem{mm}
  \textsc{Hungerbühler} (Norbert), \og A short elementary proof of the {Mohr-Mascheroni} theorem\fg, \emph{American Mathematical Monthly}, \no 101, 1994, p. 784-787.


\bibitem{Justin}
\textsc{Justin} (Jacques), \og Résolution par le pliage de l'équation du troisième degré et applications géométriques\fg, \emph{L'Ouvert - Journal de l'APMEP d'Alsace et de l'IREM de Strasbourg}, \no 42, 1986, p. 9-19. \url{https://mathinfo.unistra.fr/websites/math-info/irem/Publications/L_Ouvert/n042/o_42_9-19.pdf}

\bibitem{Lafforgue}
\textsc{Lafforgue} (Laurent), \emph{Géométrie plane et algèbre}, Paris, Hermann, 2018.

\bibitem{lang}
  \textsc{Lang} (Robert J.), \og Origami and Geometric Constructions\fg, 1996-2015, \url{http://langorigami.com/wp-content/uploads/2015/09/origami_constructions.pdf}.

\bibitem{laugwitz}
  \textsc{Laugwitz} (Detlef), \og Eine elementare {Methode} f\"{u}r {Unm\"{o}glichkeitsbeweise} bei {Konstruktionen} mit {Zirkel} und {Lineal}\fg, \emph{Elemente der Mathematik}, \no 17, 1962, p. 54-58.

\bibitem{loh1}
  \textsc{Loh} (Po-Shen), \og A Different Way to Solve Quadratic Equations\fg, 2019, \url{https://www.poshenloh.com/quadratic/}.


\bibitem{loh2}
  \textsc{Loh} (Po-Shen), \og A Simple Proof of the Quadratic Formula\fg, 2019, \url{https://arxiv.org/abs/1910.06709}.


\bibitem{manna}
  \textsc{Manna} (Zohar), \emph{Mathematical Theory of Computing}, McGraw-Hill, 1974.



\bibitem{martin}
  \textsc{Martin} (George E.), \emph{Geometric Constructions}, Springer, 1998.


\bibitem{mastin}
  \textsc{Mastin} (Luke), \og Muhammad ibn {Musa al-Khwarizmi}: {Muslim Mathematician}\fg, 2020, \url{https://www.storyofmathematics.com/islamic_alkhwarizmi.html}.


\bibitem{mccallum}
  \textsc{McCallum} (William), \og A tale of two triangles: Heron triangles and elliptic curves\fg, 2012, \url{http://blog.kleinproject.org/?p=4}.

\bibitem{mckay}
  \textsc{McKay} (Brendan D.), \og Ramsey Theory\fg, \url{http://users.cecs.anu.edu.au/~bdm/data/ramsey.html}.


\bibitem{miller}
  \textsc{Miller} (John E.), \og Langford's Problem, Remixed\fg, 2014, \url{http://dialectrix.com/langford.html}.



\bibitem{newton}
  \textsc{Newton} (Liz), \og The Power of Origami\fg, \url{https://plus.maths.org/content/power-origami}.


\bibitem{rusty}
  \textsc{Peil} (Timothy), \og The Rusty Compass Theorem\fg, 2006, \url{https://web.archive.org/web/20200720195718/http://web.mnstate.edu/peil/geometry/C2EuclidNonEuclid/1Compass.htm}.
 




\bibitem{ramanujan1}
  \textsc{Ramanujan} (Srinivasa), \og Squaring the circle\fg, \emph{Journal of the Indian Mathematical Society}, \no V, 1913, p. 138, \url{http://ramanujan.sirinudi.org/Volumes/published/ram05.pdf}.


\bibitem{ramanujan2}
  \textsc{Ramanujan} (Srinivasa), \og Modular equations and approximations to $\pi$\fg, \emph{The Quarterly Journal of Pure and Applied Mathematics}, \no XLV, 1914, p. 350-372, \url{http://ramanujan.sirinudi.org/Volumes/published/ram06.pdf}.





\bibitem{riaz}
  \textsc{Riaz} (M.), \og Geometric solutions of algebraic equations\fg, \emph{American Mathematical Monthly}, \no 69, 1962, p. 654-658.






\bibitem{rike}
  \textsc{Rike} (Tom), \og Fermat Numbers and the Heptadecagon\fg, 2005, \url{https://mathcircle.berkeley.edu/sites/default/files/BMC6/ps0506/Heptadecagon.pdf}.















 

\bibitem{robson}
 \textsc{Robson} ( Eleanor), \og Words and pictures: New light on {Plimpton} 322\fg,   \emph{American Mathematical Monthly}, \no 109, 2002, p. 105-120.


\bibitem{ross}
\textsc{Ross} (Sheldon), \emph{Initiation aux probabilités} (trad. C. Hofer, F. Dorsaz), Lausanne, Presses Polytechniques et Universitaires Romandes, 2014.



\bibitem{schumer}
  \textsc{Schumer} (Peter), \og The {Josephus} Problem: Once More around\fg, \emph{Mathematics Magazine}, \no 75, 2002, p. 12-17.

\bibitem{Signac}
\textsc{Signac} (Laurent), \emph{Divertissements mathématiques et informatiques}, Paris, H\&K, 2011.

\bibitem{silvester}
  \textsc{Silvester} (John R.), \og Ceva = ({Menelaus})$^2$\fg, \emph{The Mathematical Gazette}, \no 500, 2000, p. 268-271.






\bibitem{sipka}
  \textsc{Sipka} (Timothy), \og Alfred {Bray Kempe's} {``Proof''} of the Four-Color Theorem\fg, \emph{Math Horizons}, \no 2, 2002, p. 21-26, \url{http://www.jstor.org/stable/25678395}.





\bibitem{stillwell}
  \textsc{Stillwell} (John), \emph{Mathematics and Its History}, Springer, 2010.




\bibitem{suzuki}
  \textsc{Suzuki} (Jeff), \og A Brief History of Impossibility\fg, \emph{Mathematics Magazine}, \no 81, 2008, p. 27-38.


%\bibitem{gantert}
%  \textsc{Gantert} (Ann Xavier), \emph{Geometry}, Perfection Learning, 2008.















\bibitem{thomas}
  \textsc{Thomas} (Robin), \og An Update on the Four-Color Theorem\fg, \emph{Notices of the AMS}, \no 45, 1998, p. 848-859, \url{http://www.ams.org/notices/199807/thomas.pdf}.



  
\bibitem{toussaint}
  \textsc{Toussaint} (Godfried), \og A New Look at {Euclid's} Second Proposition\fg, \emph{The Mathematical Intelligencer}, \no 15, 1993, p. 12-23.




    
  
 




%\bibitem{ben-ari:induction}
  %\textsc{Ben-Ari} (Mordechai), \og The Many Guises of Induction\fg, 2021, \url{https://www.weizmann.ac.il/sci-tea/benari/mathematics}.










%\bibitem{wiki:bisector}
%\textsc{Wikipedia}, \og Angle bisector theorem\fg.

\bibitem{wiki:hh-axioms}
\textsc{Wikipedia}, \og {Huzita}–{Hatori} axioms\fg.




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibitem{wiki:cardano}
\textsc{Wikipédia}, \og Équation cubique\fg.

\bibitem{wiki:quadratic}
\textsc{Wikipédia}, \og Équation du second degré\fg.


\bibitem{wiki:heptadecagon}
 \textsc{Wikipédia}, \og Heptadécagone\fg.

%\bibitem{wiki:cosine}
%\textsc{Wikipédia}, \og Loi des cosinus\fg.

\bibitem{wiki:neu}
\textsc{Wikipédia}, \og Neusis\fg.

  
\bibitem{wiki:pentagon}
\textsc{Wikipédia}, \og Pentagone\fg.

\bibitem{wiki:plimpton}
\textsc{Wikipédia}, \og Plimpton 322\fg.

\bibitem{wiki:josephus}
\textsc{Wikipédia}, \og Problème de {J}osèphe\fg.


\bibitem{wiki:quad}
\textsc{Wikipédia}, \og Quadratrice d'{Hippias}\fg.

\bibitem{wiki:sexagesimal}
\textsc{Wikipédia}, \og Système sexagésimal\fg.

%\bibitem{wiki:ceva}
%\textsc{Wikipédia}, \og Théorème de Ceva\fg.

%\bibitem{wiki:menelaus}
%\textsc{Wikipédia}, \og Théorème de Ménélaüs\fg.

%\bibitem{wiki:ptolemy}
%\textsc{Wikipédia}, \og Théorème de Ptolémée\fg.

\bibitem{wiki:five}
\textsc{Wikipédia}, \og Théorème des cinq couleurs\fg.


\bibitem{wiki:four}
\textsc{Wikipédia}, \og Théorème des quatre couleurs\fg.



    


\bibitem{wiki:tri}
\textsc{Wikipédia}, \og Trisection de l'angle\fg.



  



  

  

  











  

\end{thebibliography}